Forecasting Gamma-Ray Bursts using Gravitational Waves

### 2018-11-19 Paper/Manuscript.tex
More workon ref comments

\todo{1) See end of Sec. IV on page 13, 2) Finish new Sec VIII on BH-BH inspirals}
\end{abstract}
\maketitle

@@

Here, we are interested  in exploring the predictive power of near-future ground-based GW detectors. 
%thus leave the task of reducing the latency to the experts.
More specifically, we wish to find out how much early warning future GW detectors can provide us before the merger/GRB.
Ref.~\cite{Cannon:2011vi} approached this from an algorithmic perspective by developing computationally inexpensive methods to detect inspiral
signals in GW data to provide early-warning triggers \emph{before} the merger (also see Ref.~\cite{Knowles:2018hqq} for a very recent improvement to what is currently being
used in the ALV pipeline).
We instead focus on the forecasting capabilities of the ground-based interferometers at their \emph{full design} sensitivity by computing the time interval
between the merger and the instant of detection defined in terms of a certain detection criterion which we describe below.
To this end, we consider the ALV network of 2020s composed of three L-shaped interferometers operating at their design sensitivities
and Einstein Telescope of 2030s made up of V-shaped interferometers.
Ref.~\cite{Nissanke:2012dj} investigated the capabilities of three to five-interferometer detector networks of 2020s in terms of their ranges
to inspiralling BNSs and their sky-localization of these sources. Our work here is somewhat complementary in the sense that we are mostly concerned with
how early the GW networks can detect the inspiral before the merger.

@@

Let us conclude this subsection with a brief discussion on 
%our particular choice to represent the sensitivities of the near-future ALV network.
an alternate approach to construct a model for the near-future ALV network.
This involves selecting one IFO as the ``base'' detector, e.g., LIGO Livingstone then running Monte-Carlo simulations over the four random angles $\{\theta,\phi,\psi,\iota\}$. 
We would subsequently compute the quality factor of the base IFO using the
simulation results for the four angles.
We can then obtain the quality factors of the remaining two IFOs 
--- in this particular case, LIGO Hanford and Virgo --- via their positions and orientations with respect to the base IFO. Given that the four angles
are completely random, there should be no preferred point in the angle parameter space, thus the result of the Monte-Carlo runs should converge
to the RMS average. Correspondingly, we would expect one of the other two IFOs to have somewhat higher sensitivity, and the other one to have lower sensitivity.
This network can be further expanded using exact positions of LIGO India
and KAGRA detectors. The expansion of the detector network from three to five IFOs will offer significant improvements in source localization.
However, given that the total network SNR is a sum of quadratures, the 
expanded network will not accumulate significantly higher SNR than
our model ALV network. As such our model network should serve as a good proxy for the advance warning time computations.

@@

Assuming low-spin priors  ($\chi \le 0.05$, see point 7 of Sec.~\ref{sec:idealizations}), the following values were inferred for the total mass, the chirp mass,
and the luminosity distance of the source with 90\% confidence:

@@

M = 2.74^{+0.04}_{-0.01} M_\odot,\quad M_c = 1.188^{+0.004}_{-0.002} M_\odot,\quad D= 40^{+8}_{-14}\,\text{Mpc}. \label{eq:GW170817_params}

@@

due to the fact that it can be extracted to high precision over thousands of GW cycles via Eq.~(\ref{eq:fdot}). In fact, detailed follow-up analyses (\cite{GW170817_2018}) have recently restricted the above-values to

@@

$ \left\{2.73^{+0.04}_{-0.01} M_\odot, 1.186^{+0.001}_{-0.001}M_\odot , 40.7^{+2.36}_{-2.36}\text{Mpc}\right\}$, respectively,

@@

where the errors in $D$ are at $1\sigma$ level.

How well does the ``naive'' Newtonian evolution based on the quadrupole formula perform? 
To test this, let us first compute the inspiral time, $\tau_\text{insp}$, and the number of GW cycles, $\mathcal{N}_\text{cyc}$, for this BNS system with initial frequency $f_0=30\,$Hz and final frequency $f_\text{ISCO} \simeq 1605\,$Hz. 
Eqs.~(\ref{eq:tau_insp}) and (\ref{eq:Ncyc1}) provide us with the leading-order values
%%
%
\be
\tau_\text{insp}^{0\text{PN}} \simeq 55.91\,\text{seconds}, \quad \mathcal{N}_\text{cyc}^{0\text{PN}} \simeq 2680\,\label{eq:GW170817_values_0PN}. 
\ee
%
%%
If we wish to do better, we can add all of the known post-Newtonian contributions 
up to 3.5PN\footnote{Although the equations of motion for circular orbits is now known up to 4PN \cite{4PN_EoM_paper}, we are not aware of any publications
explicitly showing the 4PN contributions to $\tau_\text{insp}$ and $\mathcal{N}_\text{cyc}$. Be that as it may, we suspect that a few PN experts have already obtained these.} resulting in
%%
%
\be
\tau_\text{insp}^{3.5\text{PN}} \simeq 56.65\,\text{seconds}, \quad \mathcal{N}_\text{cyc}^{3.5\text{PN}} \simeq 2711 \label{eq:GW170817_values_3p5PN}\, ,
\ee
%
%%
which more or less match the values above. %of $\tau_\text{insp}\approx 60$\,seconds and $\mathcal{N}_\text{cyc} \approx 3000$ which we can obtain if we use $f_i \approx 28-29\,$Hz.
%Our PN-only computation underestimates $\mathcal{N}_\text{cyc}$ because we artificially end the inspiral at the Schwarzschild ISCO whereas
%numerical simulations indicate that the inspiral carries on past the Schwarzschild ISCO separation of $6 G M/c^2$. 
%Reusing Eq.~(\ref{eq:Ncyc1}) from 1600 Hz to 2000 Hz only adds {\bf one} cycle.

@@

@@

\begin{table}[h]
\centering
\begin{tabular}{lccccccc}
%toprule
\hline\hline
$D\,$(Mpc) & \multicolumn{3}{c}{ET-B} &  & \multicolumn{3}{c}{ET-C}\T\B\\
\hline
{}& $\bar{f}_\text{ET}\,$(Hz) & \ \hspace{7mm} $T_\text{AW}$ \ \hspace{7mm} & $\bar{\rho}_{F}$ &{} & $\bar{f}_\text{ET}\,$(Hz) & \ \hspace{5mm} $T_\text{AW}\hspace{12mm}$& $\bar{\rho}_{F}$\T\B\\
%midrule
100 & $\approx\,$6.72 &  47.0\,minutes & 306 &{\qquad} & $\approx\,$3.27 & 5.34\,hours\ & 365\T\\
200 & $\approx\,$11.2 & 11.6\,minutes & 152 &{\qquad} & $\approx\,$4.10 & 2.87\,hours\ & 182 \\
400 & $\approx\,$18.2 & 3.00\,minutes & 75.7 &{\qquad} & $\approx\,$5.06 & 1.51\,hours\ & 90.5\\
1000 & $\approx\,$41.3 &17.2\,seconds & 29.8& \qquad &    $\approx\,$6.76 & 35.6\,minutes & 35.6  \\
% OLD version
%100 & 6.7237 &  2818 & 306.0 &{\qquad} & 3.2692 & 19212 & 365.4\T\\
%200 & 11.2104 & 698.6 & 152.4 &{\qquad} & 4.1028 & 10138 & 182.1 \\
%400 & 18.2234 & 179.8 & 75.74 &{\qquad} & 5.0607 & 5437 & 90.48\\
%1000 & 41.3537 &17.2 & 29.79& \qquad & 6.7553 & 2138 & 35.60  \\
\hline\hline
%bottomrule
\end{tabular}
\caption{Forecasting capabilities of Einstein Telescope. 
%We can expect over \emph{five} hours of early warning for nearby sources. 
ET-B and ET-C refer to the different configurations shown in Fig.~\ref{fig:ETB2030}. For the advance warning times, we only present the result of the more accurate 3.5PN computation. $\bar{f}_\text{ET}$ is the threshold frequency at which
ET-B/C accumulate SNR of 15 which we take to be our detection criterion. Note that both $T_\text{AW}$ and $\bar\rho_F$ are larger for ET-C
due to its improved sensitivity in the $1\,\text{Hz}\lesssim f\lesssim 30\,$Hz regime compared to ET-B as is clear in Fig.~\ref{fig:ETB2030}.
These results and those of Table.~\ref{table:LIGO2020} are summarized in Fig.~\ref{fig:summary}.}\label{table:ET}
\end{table}
%
%
As the results of Table~\ref{table:ET} indicate, ET-B will provide $\gtrsim 10\,$ minutes of warning time for BNS inspirals at $D\lesssim 200\,$Mpc.
Instead of speculating whether or not this is enough to forecast GRBs, we turn our attention to ET-C's forecasting capabilities.
As can be seen from Fig.~\ref{fig:ETB2030}, ET-C's increased sensitivity in the $1\,\text{Hz}\lesssim f\lesssim 30\,$Hz regime will significantly lengthen the advance warning times.
This is best highlighted by the sixth colum of Table~\ref{table:ET}, 
where we show that ET-C will provide up to \emph{five hours} of early warning time for nearby ($D\lesssim 100\,$Mpc) neutron star mergers.
Moreover, BNS inspirals as far as 200\,Mpc (the ALV network's range) will persist in the ET-C sensitivity band for a few hours after their initial detection ($T_\text{AW} > 2\,$ hours).
Current BNS merger rate, inferred from A-LIGO's O1-O2 periods, implies that $\sim 40$ to $\sim600$ sources within a volume of $(200\,\text{Mpc})^3$ will be detected yearly by ET-C.

@@

inside a radius of $D\approx 1000\,$Mpc to within $ 10\,\text{deg}^2$ and more than half the sources within $D\lesssim 400\,$Mpc to $\approx 1\,\text{deg}^2$ \cite{Mills:2017urp}.
Let us reemphasize that such good localization requires at least a second detector with comparable sensitivity to ET as ET alone has rather poor localization.
Erring on the side of optimism, we conclude that the 2030s hold in store real-time observations of dozens to hundreds of gamma-ray bursts
\footnote{As the ET construction/commissioning timeline is currently uncertain, we can only speculate that ET-C will be operational after 2030,
which is roughly in line with the milli-Hertz space interferometer LISA's timeline \cite{Audley:2017drz}.}.

>>>>

\todo{1) See end of Sec. IV on page 13, 2) Finish new Sec VIII on BH-BH inspirals, 3) page 19. footnote}
\end{abstract}
\maketitle

@@

Here, we are interested  in exploring the predictive power of near-future ground-based GW detectors. 
%thus leave the task of reducing the latency to the experts.
More specifically, we wish to find out how much early warning future GW detectors can provide us before the merger/GRB.
Ref.~\cite{Cannon:2011vi} approached this from an algorithmic perspective by developing computationally inexpensive methods to detect inspiral
signals in GW data to provide early-warning triggers \emph{before} the merger (also see Ref.~\cite{Knowles:2018hqq} for a very recent improvement to what is currently being
used in the ALV pipeline).
We instead focus on the forecasting capabilities of the ground-based interferometers at their \emph{full design} sensitivity by computing the time interval
between the merger and the instant of detection defined in terms of a certain detection criterion which we describe below.
To this end, we consider the ALV network of 2020s composed of three L-shaped interferometers operating at their design sensitivities
and Einstein Telescope of 2030s made up of V-shaped interferometers.
Ref.~\cite{Nissanke:2012dj} investigated the capabilities of three to five-interferometer detector networks of 2020s in terms of their ranges
to inspiralling BNSs and their sky-localization of these sources. 
More recently, Ref.~\cite{Chan:2018csa} added to this results pertaining to the future networks of 2030s both in terms of source localization and advance
warning prospects.
Our work here is complementary to these articles in the sense that we focus on
the details of how to quantify the advance warning.
%we are mostly concerned with how early the GW networks can detect the inspiral before the merger.

@@

Let us conclude this subsection with a brief discussion on 
%our particular choice to represent the sensitivities of the near-future ALV network.
an alternate approach to construct a model for the near-future ALV network.
This involves selecting one IFO as the ``base'' detector, e.g., LIGO Livingstone then running Monte-Carlo simulations over the four random angles $\{\theta,\phi,\psi,\iota\}$. 
We would subsequently compute the quality factor of the base IFO using the
simulation results for the four angles.
We can then obtain the quality factors of the remaining two IFOs 
--- in this particular case, LIGO Hanford and Virgo --- via their positions and orientations with respect to the base IFO. Given that the four angles
are completely random, there should be no preferred point in the angle parameter space, thus the result of the Monte-Carlo runs should more or less agree with the RMS average. Correspondingly, we would expect one of the other two IFOs to have somewhat higher sensitivity, and the other one to have lower sensitivity.
This network can be further expanded using exact positions of LIGO India
and KAGRA detectors. The expansion of the detector network from three to five IFOs will offer significant improvements in source localization.
However, given that the total network SNR is a sum of quadratures, the 
expanded network will not accumulate significantly higher SNR than
our model ALV network. As such our model network should serve as a good proxy for the advance warning time computations.

@@

Assuming low-spin priors  ($\chi \le 0.05$, see point 7 of Sec.~\ref{sec:idealizations}), the following updated values were inferred for the total mass, the chirp mass,
and the luminosity distance of the source with 90\% confidence \cite{GW170817_2018}:

@@

M = 2.73^{+0.04}_{-0.01} M_\odot,\quad M_c = 1.186^{+0.001}_{-0.001} M_\odot,\quad D= 40.7^{+2.36}_{-2.36}\,\text{Mpc}. \label{eq:GW170817_params}

@@

due to the fact that it can be extracted to high precision over thousands of GW cycles via Eq.~(\ref{eq:fdot}). 
%In fact, detailed follow-up analyses () have recently restricted the above-values to

@@

%$ \left\{2.73^{+0.04}_{-0.01} M_\odot, 1.186^{+0.001}_{-0.001}M_\odot , 40.7^{+2.36}_{-2.36}\text{Mpc}\right\}$, respectively,

@@

%where the errors in $D$ are at $1\sigma$ level.

How well does the ``naive'' Newtonian evolution based on the quadrupole formula perform? 
To test this, let us first compute the inspiral time, $\tau_\text{insp}$, and the number of GW cycles, $\mathcal{N}_\text{cyc}$, for this BNS system with initial frequency $f_0=30\,$Hz and final frequency $f_\text{ISCO} \simeq 1605\,$Hz. 
Eqs.~(\ref{eq:tau_insp}) and (\ref{eq:Ncyc1}) provide us with the leading-order values
%%
%
\be
\tau_\text{insp}^{0\text{PN}} \simeq 56.06\,\text{seconds}, \quad \mathcal{N}_\text{cyc}^{0\text{PN}} \simeq 2687\,\label{eq:GW170817_values_0PN}. 
\ee
%
%%
If we wish to do better, we can add all the post-Newtonian contributions up to 2PN resulting in
%%
%
\be
\tau_\text{insp}^{2\text{PN}} \simeq 57.06\,\text{seconds}, \quad \mathcal{N}_\text{cyc}^{2\text{PN}} \simeq 2735 \label{eq:GW170817_values_3p5PN}\, ,
\ee
%
%%
which more or less match the detected inspiral time and number of cycles. We stop at 2PN because the currently available 3.5PN expressions require the value of $\nu$
which is not well known for GW170817.
%
%\footnote{Although the equations of motion for circular orbits is now known up to 4PN \cite{4PN_EoM_paper}, we are not aware of any publications
%explicitly showing the 4PN contributions to $\tau_\text{insp}$ and $\mathcal{N}_\text{cyc}$. Be that as it may, we suspect that a few PN experts have already obtained these.} 
%of $\tau_\text{insp}\approx 60$\,seconds and $\mathcal{N}_\text{cyc} \approx 3000$ which we can obtain if we use $f_i \approx 28-29\,$Hz.
%Our PN-only computation underestimates $\mathcal{N}_\text{cyc}$ because we artificially end the inspiral at the Schwarzschild ISCO whereas
%numerical simulations indicate that the inspiral carries on past the Schwarzschild ISCO separation of $6 G M/c^2$. 
%Reusing Eq.~(\ref{eq:Ncyc1}) from 1600 Hz to 2000 Hz only adds {\bf one} cycle.

@@

In mid 2020s LIGO will first be upgraded to A+ then to Voyager. 
However, both versions will have comparable sensitivities to Advanced LIGO
at low frequencies (see Fig.~1 of Ref.~\cite{Mills:2017urp}).
Therefore, they will not start accumulating significant SNR until $f\approx 10\,$Hz, thus offering us gains of mere minutes compared to $T_\text{AW}$ in Table~\ref{table:LIGO2020}. For this reason, we move on to 2030s: the 
era of Einstein Telescope.

@@

\begin{table}[h]
\centering
\begin{tabular}{lccccccc}
%toprule
\hline\hline
$D\,$(Mpc) & \multicolumn{3}{c}{ET-B} &  & \multicolumn{3}{c}{ET-C}\T\B\\
\hline
{}& $\bar{f}_\text{ET}\,$(Hz) & \ \hspace{7mm} $T_\text{AW}$ \ \hspace{7mm} & $\bar{\rho}_{F}$ &{} & $\bar{f}_\text{ET}\,$(Hz) & \ \hspace{5mm} $T_\text{AW}\hspace{12mm}$& $\bar{\rho}_{F}$\T\B\\
%midrule
100 & $\approx\,$6.72 &  47.0\,minutes & 306 &{\qquad} & $\approx\,$3.27 & 5.34\,hours\ & 365\T\\
200 & $\approx\,$11.2 & 11.6\,minutes & 152 &{\qquad} & $\approx\,$4.10 & 2.87\,hours\ & 182 \\
400 & $\approx\,$18.2 & 3.00\,minutes & 75.7 &{\qquad} & $\approx\,$5.06 & 1.51\,hours\ & 90.5\\
1000 & $\approx\,$41.3 &17.2\,seconds & 29.8& \qquad &    $\approx\,$6.76 & 35.6\,minutes & 35.6  \\
% OLD version
%100 & 6.7237 &  2818 & 306.0 &{\qquad} & 3.2692 & 19212 & 365.4\T\\
%200 & 11.2104 & 698.6 & 152.4 &{\qquad} & 4.1028 & 10138 & 182.1 \\
%400 & 18.2234 & 179.8 & 75.74 &{\qquad} & 5.0607 & 5437 & 90.48\\
%1000 & 41.3537 &17.2 & 29.79& \qquad & 6.7553 & 2138 & 35.60  \\
\hline\hline
%bottomrule
\end{tabular}
\caption{Forecasting capabilities of Einstein Telescope. 
%We can expect over \emph{five} hours of early warning for nearby sources. 
ET-B and ET-C refer to the different configurations shown in Fig.~\ref{fig:ETB2030}. For the advance warning times, we only present the result of the more accurate 3.5PN computation. $\bar{f}_\text{ET}$ is the threshold frequency at which
ET-B/C accumulate SNR of 15 which we take to be our detection criterion. Note that both $T_\text{AW}$ and $\bar\rho_F$ are larger for ET-C
due to its improved sensitivity in the $1\,\text{Hz}\lesssim f\lesssim 30\,$Hz regime compared to ET-B as is clear in Fig.~\ref{fig:ETB2030}.
These results and those of Table.~\ref{table:LIGO2020} are summarized in Fig.~\ref{fig:summary}.}\label{table:ET}
\end{table}
%
%
As the results of Table~\ref{table:ET} indicate, ET-B will provide $\gtrsim 10\,$ minutes of warning time for BNS inspirals at $D\lesssim 200\,$Mpc.
Instead of speculating whether or not this is enough to forecast GRBs, we turn our attention to ET-C's forecasting capabilities.
As can be seen from Fig.~\ref{fig:ETB2030}, ET-C's increased sensitivity in the $1\,\text{Hz}\lesssim f\lesssim 30\,$Hz regime will significantly lengthen the advance warning times.
This is best highlighted by the sixth colum of Table~\ref{table:ET}, 
where we show that ET-C will provide up to \emph{five hours} of early warning time for nearby ($D\lesssim 100\,$Mpc) neutron star mergers.
Moreover, BNS inspirals as far as 200\,Mpc (the ALV network's range) will persist in the ET-C sensitivity band for a few hours after their initial detection ($T_\text{AW} > 2\,$ hours).
Current BNS merger rate, inferred from A-LIGO's O1-O2 periods, implies that $\sim 40$ to $\sim600$ sources within a volume of $(200\,\text{Mpc})^3$ will be detected yearly by ET-C. 
Our values for $T_\text{AW}$ for 200 and 400\,Mpc are roughly consistent
with the findings of Ref.~\cite{Chan:2018csa}.

@@

inside a radius of $D\approx 1000\,$Mpc to within $ 10\,\text{deg}^2$ \cite{Mills:2017urp}, but mostly after the merger.
%and more than half the sources within $D\lesssim 400\,$Mpc to $\approx 1\,\text{deg}^2$ . 
On the other hand, ET with CE together will localize 
$\sim 80, 20\%$ of the BNSs
within 200, 400\,Mpc to $1\,$deg$^2$, respectively \cite{Chan:2018csa},
%Let us reemphasize that such good localization requires at least a second detector %with comparable sensitivity to ET as ET alone has rather poor localization.
since CE will be sensitive enough to start accumulating SNR at $f\approx 5\,$Hz from BNSs within 400\,Mpc corresponding to $T_\text{AW}\gtrsim 1.5\,$hours.
Even without the help of a second detector, it will be possible
to localize $\sim 5\%$ of these transients to within $10\,$deg$^2$ using
the Earth's rotation because these nearby sources will spend hours in the ET-C band \cite{Zhao:2017cbb}.
This percentage amounts to $\sim 5$ events per year.
%Additionally, these localizations are mostly post-merger as neither Voyager,
Erring on the side of optimism, we conclude that the 2030s hold in store real-time observations of dozens to hundreds of gamma-ray bursts and the subsequent kilonovae
\footnote{As the ET construction/commissioning timeline is currently uncertain, we can only speculate that ET-C will be operational in mid 2030s,
which is roughly in line with the timeline of the milli-Hertz space interferometer LISA \cite{Audley:2017drz} and may be five to ten years ahead of CE \todo{citation}}.
